---
title: "EDA - R Sangole"
output:
  html_notebook:
    highlight: zenburn
    theme: flatly
    toc: yes
  pdf_document:
    fig_width: 9
    highlight: zenburn
    toc: yes
---
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(glue)
library(janitor)
library(caret)
library(lattice)
library(skimr)
library(latticeExtra)
library(lattice)
library(plotly)
library(ggthemr)
library(GGally)
ggthemr('light')
```
```{r load_data}
load('../cache/raw_train.rdata')
```

# Hypotheses

## H1 - Some cover types have large h distances

- Some covertypes have very large absolute values of distances. cottonwood has < 550. 
- This could be an additional indicator variable. "h_dist_gt_500"
- Some binning and indicator variables might show promise
- Therea are also some linear structures in the data that perhaps can be exploited

```{r}
raw_train %>%    
    dplyr::select(cover_type, contains("hydrology")) %>%
    ggplot(aes(horizontal_distance_to_hydrology, vertical_distance_to_hydrology)) +
    geom_point(aes(color = cover_type), alpha = 0.1, stroke = 0) +
    geom_smooth() +
    facet_wrap(~cover_type) +
    theme(legend.position="none")
```

```{r}
raw_train %>%    
    ggplot(aes(cover_type, horizontal_distance_to_hydrology))+
    geom_jitter(alpha=0.05)+
    geom_boxplot(fill=NA)+
    geom_hline(yintercept = 0, color='gray')
```

. These plots show h_dist is either exactly 0 or there's a positive number

```{r}
densityplot(~horizontal_distance_to_hydrology|cover_type,raw_train, plot.points=F,bw = 10)
```

```{r}
raw_train %>%
    group_by(cover_type) %>% 
    summarise(max = max(horizontal_distance_to_hydrology), min = min(horizontal_distance_to_hydrology))
```


## H2 - Some cover types have only positive v distances

- There seems to be some truth in this.
- cottonwood is absolutely almost around 0
- spruce and lodgepole have strong -ve components but not so much the other trees. 
- perhaps an indicator variable would improve the model "v_dist_lt_0"
- also, some v_dsts are capped at ~270 (like ponderosa or aspen or douglasfir). Only lodgepole goes > 450. this is a definite predictor. we should have an indicator var like "v_dist_gt_450". 
- come to think of it... perhaps a binning strategy would benefit here, driving better predictive performance than nominal variable

```{r}
raw_train %>%    
    dplyr::select(cover_type, contains("hydrology")) %>%
    ggplot(aes(cover_type, vertical_distance_to_hydrology))+
    geom_jitter(alpha=0.05)+
    geom_hline(yintercept = 0, color='gray')
```
```{r}
raw_train %>%
    group_by(cover_type) %>% 
    summarise(max = max(vertical_distance_to_hydrology), min = min(vertical_distance_to_hydrology))
```

## H3 - Direct distances add value

- Not sure that it does... although it might clean up the distributions a bit. we'll have to experiment

```{r}
raw_train <- raw_train %>% 
    mutate(dist = sqrt(horizontal_distance_to_hydrology^2+vertical_distance_to_hydrology^2))
raw_train %>% 
    ggplot(aes(cover_type, dist)) +
    geom_jitter(alpha = 0.1, stroke = 0) +
    geom_boxplot(fill=NA)
```


```{r message=FALSE, warning=FALSE, fig.height=6, fig.width=10}
raw_train %>%    
    dplyr::select(cover_type, contains("hill")) %>%
    ggpairs(aes(color=cover_type), alpha=0.2)
```

# EDA Takeaways & Ideas

## Observations

1. {lodgepole:aspen}, {ponderosa:cottonwood} have similar distributions across soils
1. no strong correlations among the soils
1. clear distinctions among wilderness - {{spruce:lodgepole}:{aspen:krummholz}}, {ponderosa:douglasfir}, {cottonwood}
1. Some strong interactions between wilderness & soil type and cover_type
1. krummholz is strongly soils > 35
1. Shade measures (esp9am) seem quite predictive

## Ideas

1. Interactions can be modeled explicitely for linear models or to boost accuracy of the heuristic models, especially some wilderness*soiltypes
1. change hori and vertical distance to hydrology to a straight line distance
1. lots of opportunities for indicator variables after binning continuous variables
1. PCA to dimension reduce soil / hillside vars?
